{Good covers and algebras on conically smooth spaces}

### 2025-09-11 paperversion/paper.tex
unzip along S_X

\newcommand{\R}{\mbb{R}}
\DeclareMathOperator{\Imm}{Im}
\newcommand{\id}{\mathrm{id}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\dfnt}{\bfseries \itshape}
\newcommand{\hyph}{{\mbox{-}}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\Fun}{\mathrm{Fun}}
\newcommand{\Tangent}{\mathrm{T}}
\newcommand{\Normal}{\mathrm{N}}
\newcommand{\evv}{\mathrm{ev}}  
\DeclareMathOperator{\Unzip}{Unzip}
\DeclareMathOperator{\Link}{Link}
\DeclareMathOperator{\Span}{span}
\newcommand{\opp}{{\mathrm{op}}}

@@

\begin{lemma}
    For every CSS $X$, the inclusion $S_X\hookrightarrow X$ is a proper constructible embedding of conically smooth spaces.
\end{lemma}
\begin{proof}
    Let $X\to P$ be the stratification of $X$ and let $X\to P\to\mbb{P}=\mbb{Z}^{\opp}\times\mbb{Z}^{\opp}$ be its depth-dimension stratification \cite[Lemma 2.4.10]{ayala2017local}. The inclusion $\mbb{Z}_{>0}^{\opp}\subset\mbb{Z}^\opp$ (where in the subscript $>$ refers to the ordinary ordering and not to its opposite) is a full subcategory inclusion and is thus consecutive \cite[Definition 2.3.1]{ayala2017local}, hence so is $\mbb{P}_{+}=\mbb{Z}_{>0}^\opp\times\mbb{Z}^\opp\hookrightarrow\mbb{P}$. Since consecutive poset maps are stable under pullbacks, $\mbb{P}_{+}\times_{\mbb{P}}P\hookrightarrow P$ is likewise consecutive, proving that the inclusion $S_X=X|_{\mbb{P}_+\times_{\mbb{P}}P}\hookrightarrow X$ is a constructible map of conically smooth spaces \cite[Lemma 3.4.5, Example 3.4.7]{ayala2017local}. Finally, as $\mbb{Z}_{>0}^{\opp}\subset\mbb{Z}^\opp$ is downward-closed, $S_X$ is closed in $X$, so (since CSSs are Hausdorff) the inclusion map is proper.
\end{proof}

\begin{proposition}\label{7GBVGBA}
    $S_X\subset X$ proper constructible bundle with $\Unzip_{S_X}(X)\cong\Unzip(X)$ and $\Link_{S_X}(X)=\partial\Unzip(X)$.
\end{proposition}

>>>>

\newcommand{\R}{\mbb{R}}
\DeclareMathOperator{\Imm}{Im}
\newcommand{\id}{\mathrm{id}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\dfnt}{\bfseries \itshape}
\newcommand{\hyph}{{\mbox{-}}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\Fun}{\mathrm{Fun}}
\newcommand{\Tangent}{\mathrm{T}}
\newcommand{\Normal}{\mathrm{N}}
\newcommand{\evv}{\mathrm{ev}}  
\DeclareMathOperator{\Unzip}{Unzip}
\DeclareMathOperator{\Link}{Link}
\DeclareMathOperator{\Span}{span}
\newcommand{\opp}{{\mathrm{op}}}
\DeclareMathOperator{\depth}{depth}

@@

\begin{lemma}\label{8W651JS}
    For every CSS $X$, the inclusion $S_X\hookrightarrow X$ is a proper constructible embedding of conically smooth spaces.
\end{lemma}
\begin{proof}
    Let $X\to P$ be the stratification of $X$ and let $X\to P\to\mbb{P}=\mbb{Z}^{\opp}\times\mbb{Z}^{\opp}$ be its depth-dimension stratification \cite[Lemma 2.4.10]{ayala2017local}. The inclusion $\mbb{Z}_{>0}^{\opp}\subset\mbb{Z}^\opp$ (where in the subscript $>$ refers to the ordinary ordering and not to its opposite) is a full subcategory inclusion and is thus consecutive \cite[Definition 2.3.1]{ayala2017local}, hence so is $\mbb{P}_{+}=\mbb{Z}_{>0}^\opp\times\mbb{Z}^\opp\hookrightarrow\mbb{P}$. Since consecutive poset maps are stable under pullbacks, $\mbb{P}_{+}\times_{\mbb{P}}P\hookrightarrow P$ is likewise consecutive, proving that the inclusion $S_X=X|_{\mbb{P}_+\times_{\mbb{P}}P}\hookrightarrow X$ is a constructible map of conically smooth spaces \cite[Lemma 3.4.5, Example 3.4.7]{ayala2017local}. Finally, as $\mbb{Z}_{>0}^{\opp}\subset\mbb{Z}^\opp$ is downward-closed, $S_X$ is closed in $X$, so (since CSSs are Hausdorff) the inclusion map is proper.
\end{proof}

By virtue of \Cref{8W651JS} we may apply unzip $X$ along $S_X$ according to \cite[Proposition 7.3.10]{ayala2017local}.

\begin{proposition}\label{7GBVGBA}
    The unzip of $X$ is its unzip along its singular locus. That is, $\Unzip(X)\cong\Unzip_{S_X}(X)$ and consequently $\partial\Unzip(X)\cong\Link_{S_X}(X)$.
\end{proposition}
\begin{proof}
    If $\depth(X)=0$ the statement holds trivially, so let us suppose that $\depth(X)\leq n\geq1$ and that the statement holds for all CSSs $Y$ with $\depth(Y)\leq n-1$.
    The problem is local, so, suppressing euclidean factors for simplicity, let $X=C(Z)$ be a conically smooth basic. By \cite[Lemma 7.3.5, (5) and (6)]{ayala2017local} and by the inductive hypothesis we have $\Unzip(C(Z))=\R_{\geq0}\times\Unzip(Z)=\R_{\geq0}\times\Unzip_{S_Z}(Z)$, as $Z$ has strictly lower depth. On the other hand, since $S_{C(Z)}=C(S_Z)$ by inspection, we have $\Unzip_{S_{C(Z)}}C(Z)=\Unzip_{C(S_Z)}C(Z)=\R_{\geq0}\times\Unzip_{S_Z}(Z)$ by construction (cf.\ the proof of \cite[Proposition 7.3.10]{ayala2017local}), proving $\Unzip(C(Z))=\Unzip_{S_{C(Z)}}(C(Z))$.
\end{proof}
